---
title: "Calculate Travel Cost"
author: "Weizhe Weng"
date: "3/13/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calculate travel distance 

In this section, we calculate travel distance using OpenStreetMap,a a free and open geographic database.   

```{r}
library(sf)
library(osrm)
library(tidyverse)
library(dplyr)
```

Calculate travel distances (by driving, in meters) between central park and visitor home blockgroup for year 2019 and year 2020.

```{r}
options(osrm.server = "http://localhost:5000/",osrm.profile = "driving")
getOption("osrm.server")

#Year 2019
centralp19 <- read.csv("~/Dropbox/Research/Pandemic and Recreation/Dataset/Central Park/centralpark_visit_2019.csv")

centralp19<-centralp19%>%select(c(-unique_ID,-minutes,-meters))
  
centralp19 = centralp19 %>% mutate(unique_ID = paste(as.character(park_longitude),as.character(park_latitude),sep="_"))

dest19 <- centralp19 %>% 
  select(unique_ID,park_latitude,park_longitude) %>%
  distinct() %>% 
  relocate(park_latitude, .after = last_col()) 

home19 <- centralp19 %>%
  select(visitor_home_cbgs,visitor_home_latitude,visitor_home_longitude) %>%
  distinct() %>%
  relocate(visitor_home_latitude, .after = last_col()) %>% 
  filter(!is.na(visitor_home_latitude))

fulltd19 <- data.frame()

for (n in 1:nrow(home19)){
  singletd <- osrmTable(src = home19[n,c("visitor_home_longitude","visitor_home_latitude")],
                        dst = dest19[,c("park_longitude","park_latitude")],measure = 'distance')
  temp <- singletd$distances
  temp <-  as.data.frame(temp) %>% cbind(id = row.names(temp), .)
  fulltd19 <- rbind(fulltd19,temp)
}

fulltdl19 <- pivot_longer(fulltd19, cols = 2:33, names_to = "park_id",values_to = "driving_meters") %>%
  mutate(id = as.numeric(id))


#Year 2020
centralp20 <- read.csv("~/Dropbox/Research/Pandemic and Recreation/Dataset/Central Park/centralpark_visit_2020.csv")

centralp20<-centralp20%>%select(c(-unique_ID,-minutes,-meters))
  
centralp20 = centralp20 %>% mutate(unique_ID = paste(as.character(park_longitude),as.character(park_latitude),sep="_"))

dest20 <- centralp20 %>% 
  select(unique_ID,park_latitude,park_longitude) %>%
  distinct() %>% 
  relocate(park_latitude, .after = last_col()) 

home20 <- centralp20 %>%
  select(visitor_home_cbgs,visitor_home_latitude,visitor_home_longitude) %>%
  distinct() %>%
  relocate(visitor_home_latitude, .after = last_col()) %>% 
  filter(!is.na(visitor_home_latitude))

fulltd20 <- data.frame()

for (n in 1:nrow(home20)){
  singletd <- osrmTable(src = home20[n,c("visitor_home_longitude","visitor_home_latitude")],
                        dst = dest20[,c("park_longitude","park_latitude")],measure = 'distance')
  temp <- singletd$distances
  temp <-  as.data.frame(temp) %>% cbind(id = row.names(temp), .)
  fulltd20 <- rbind(fulltd20,temp)
}

fulltdl20 <- pivot_longer(fulltd20, cols = 2:36, names_to = "park_id",values_to = "driving_meters") %>%
  mutate(id = as.numeric(id))

centralp20 <- left_join(centralp20, fulltdl20, by = c("unique_ID" = "park_id", "visitor_home_cbgs" = "id"))


```

Calculate travel time (by driving, in minutes) between central park and visitor home blockgroup for year 2019 and year 2020.

```{r}
#year 2019
fulltm19 <- data.frame()

for (n in 1:nrow(home19)){
  singletm <- osrmTable(src = home19[n,c("visitor_home_longitude","visitor_home_latitude")],
                        dst = dest19[,c("park_longitude","park_latitude")])
  temp <- singletm$duration
  temp<- as.data.frame(temp) %>% cbind(id = row.names(temp), .)
  fulltm19 <- rbind(fulltm19,temp)
}

fulltml19 <- pivot_longer(fulltm19, cols = 2:33,names_to = "park_id",values_to = "driving_minutes") %>%
  mutate(id = as.numeric(id))

centralp19 <- left_join(centralp19, fulltml19, by = c("unique_ID" = "park_id", "visitor_home_cbgs" = "id"))

#year2020
fulltm20 <- data.frame()

for (n in 1:nrow(home20)){
  singletm <- osrmTable(src = home20[n,c("visitor_home_longitude","visitor_home_latitude")],
                        dst = dest20[,c("park_longitude","park_latitude")])
  temp <- singletm$duration
  temp<- as.data.frame(temp) %>% cbind(id = row.names(temp), .)
  fulltm20 <- rbind(fulltm20,temp)
}

fulltml20 <- pivot_longer(fulltm20, cols = 2:36,names_to = "park_id",values_to = "driving_minutes") %>%
  mutate(id = as.numeric(id))

centralp20 <- left_join(centralp20, fulltml20, by = c("unique_ID" = "park_id", "visitor_home_cbgs" = "id"))

```

Calculate travel distances (by walking, in meters) between central park and visitor home blockgroup for year 2019 and year 2020.

```{r}
options(osrm.server = "http://localhost:5000/",osrm.profile = "foot")
getOption("osrm.server")

#Year 2019

walkfulltd19 <- data.frame()

for (n in 1:nrow(home19)){
  singletd <- osrmTable(src = home19[n,c("visitor_home_longitude","visitor_home_latitude")],
                        dst = dest19[,c("park_longitude","park_latitude")],measure = 'distance')
  temp <- singletd$distances
  temp <-  as.data.frame(temp) %>% cbind(id = row.names(temp), .)
  walkfulltd19 <- rbind(walkfulltd19,temp)
}

walkfulltdl19 <- pivot_longer(walkfulltd19, cols = 2:33, names_to = "park_id",values_to = "walking_meters") %>%
  mutate(id = as.numeric(id))

centralp19 <- left_join(centralp19, walkfulltdl19, by = c("unique_ID" = "park_id", "visitor_home_cbgs" = "id"))

#Year 2020

walkfulltd20 <- data.frame()

for (n in 1:nrow(home20)){
  singletd <- osrmTable(src = home20[n,c("visitor_home_longitude","visitor_home_latitude")],
                        dst = dest20[,c("park_longitude","park_latitude")],measure = 'distance')
  temp <- singletd$distances
  temp <-  as.data.frame(temp) %>% cbind(id = row.names(temp), .)
  walkfulltd20 <- rbind(walkfulltd20,temp)
}

walkfulltdl20 <- pivot_longer(walkfulltd20, cols = 2:36, names_to = "park_id",values_to = "walking_meters") %>%
  mutate(id = as.numeric(id))

centralp20 <- left_join(centralp20, walkfulltdl20, by = c("unique_ID" = "park_id", "visitor_home_cbgs" = "id"))


```

Calculate travel time (by walking, in minutes) between central park and visitor home blockgroup for year 2019 and year 2020.

```{r}
#year 2019
walkfulltm19 <- data.frame()

for (n in 1:nrow(home19)){
  singletm <- osrmTable(src = home19[n,c("visitor_home_longitude","visitor_home_latitude")],
                        dst = dest19[,c("park_longitude","park_latitude")])
  temp <- singletm$duration
  temp<- as.data.frame(temp) %>% cbind(id = row.names(temp), .)
  walkfulltm19 <- rbind(walkfulltm19,temp)
}

walkfulltml19 <- pivot_longer(walkfulltm19, cols = 2:33,names_to = "park_id",values_to = "walking_minutes") %>%
  mutate(id = as.numeric(id))

centralp19 <- left_join(centralp19, walkfulltml19, by = c("unique_ID" = "park_id", "visitor_home_cbgs" = "id"))

#year2020
walkfulltm20 <- data.frame()

for (n in 1:nrow(home20)){
  singletm <- osrmTable(src = home20[n,c("visitor_home_longitude","visitor_home_latitude")],
                        dst = dest20[,c("park_longitude","park_latitude")])
  temp <- singletm$duration
  temp<- as.data.frame(temp) %>% cbind(id = row.names(temp), .)
  walkfulltm20 <- rbind(walkfulltm20,temp)
}

walkfulltml20 <- pivot_longer(walkfulltm20, cols = 2:36,names_to = "park_id",values_to = "walking_minutes") %>%
  mutate(id = as.numeric(id))

centralp20 <- left_join(centralp20, walkfulltml20, by = c("unique_ID" = "park_id", "visitor_home_cbgs" = "id"))

```

Combine walking and driving mode using a threshold of 800 meter. In this case, if the walking distance is less than 800 meter, people will walk, if the walking distance is larger than 800 meter, people with drive. 

```{r}
#Select data based on walking distance
walkd19<-walkfulltdl19%>%filter(walking_meters<800)
walkd19$home_park_id<-paste(walkd19$id, walkd19$park_id, sep = "-")

walkd20<-walkfulltdl20%>%filter(walking_meters<800)
walkd20$home_park_id<-paste(walkd20$id, walkd20$park_id, sep = "-")

#exclude data in driving mode based on id and park id
fulltdl19$home_park_id<-paste(fulltdl19$id, fulltdl19$park_id, sep = "-")
unique_ids19<-unique(walkd19$home_park_id)
fulltdwl19<-fulltdl19%>%filter(!(home_park_id %in% unique_ids19))

fulltdl20$home_park_id<-paste(fulltdl20$id, fulltdl20$park_id, sep = "-")
unique_ids20<-unique(walkd20$home_park_id)
fulltdwl20<-fulltdl20%>%filter(!(home_park_id %in% unique_ids20))

```

Merge travel distance and travel time
```{r}
#walking
walkfulltml19$home_park_id<-paste(walkfulltml19$id, walkfulltml19$park_id, sep = "-")
walkfulltml20$home_park_id<-paste(walkfulltml20$id, walkfulltml20$park_id, sep = "-")

#driving
fulltml19$home_park_id<-paste(fulltml19$id, fulltml19$park_id, sep = "-")
fulltml20$home_park_id<-paste(fulltml20$id, fulltml20$park_id, sep = "-")

#merge based on home_park_id
walk_19<-left_join(walkd19,walkfulltml19,by="home_park_id")
walk_19$mode<-"walk"
walk_20<-left_join(walkd20,walkfulltml20,by="home_park_id")
walk_20$mode<-"walk"

drive_19<-left_join(fulltdl19,fulltml19,by="home_park_id")
drive_19$mode<-"drive"
drive_20<-left_join(fulltdl20,fulltml20,by="home_park_id")
drive_20$mode<-"drive"

walk_19<-walk_19%>% rename("homeid" = "id.x")%>% rename("parkid" = "park_id.x")%>% rename("travel_meters" = "walking_meters")%>% rename("travel_minutes" = "walking_minutes")

walk_19<-walk_19%>%select(-c("id.y","park_id.y"))

walk_20<-walk_20%>%rename("homeid" = "id.x")%>% rename("parkid" = "park_id.x")%>% rename("travel_meters" = "walking_meters")%>% rename("travel_minutes" = "walking_minutes")

walk_20<-walk_20%>%select(-c("id.y","park_id.y"))

drive_19<-drive_19%>% rename("homeid" = "id.x")%>% rename("parkid" = "park_id.x")%>% rename("travel_meters" = "driving_meters")%>% rename("travel_minutes" = "driving_minutes")

drive_19<-drive_19%>%select(-c("id.y","park_id.y"))

drive_20<-drive_20%>% rename("homeid" = "id.x")%>% rename("parkid" = "park_id.x")%>% rename("travel_meters" = "driving_meters")%>% rename("travel_minutes" = "driving_minutes")

drive_20<-drive_20%>%select(-c("id.y","park_id.y"))

travel_19<-rbind(walk_19,drive_19)

travel_20<-rbind(walk_20,drive_20)

dest19$parkid<-row.names(dest19)
home19$homeid<-row.names(home19)

travel_19<-left_join(travel_19,dest19,by="parkid")
travel_19$homeid<-as.character(travel_19$homeid)
travel_19<-left_join(travel_19,home19,by="homeid")

dest20$parkid<-row.names(dest20)
home20$homeid<-row.names(home20)

travel_20<-left_join(travel_20,dest20,by="parkid")
travel_20$homeid<-as.character(travel_20$homeid)
travel_20<-left_join(travel_20,home20,by="homeid")

travel_19$home_park_idn<-paste(travel_19$visitor_home_cbgs, travel_19$unique_ID, sep = "-")

travel_20$home_park_idn<-paste(travel_20$visitor_home_cbgs, travel_20$unique_ID, sep = "-")

centralp19$home_park_idn<-paste(centralp19$visitor_home_cbgs, centralp19$unique_ID, sep = "-")

centralp20$home_park_idn<-paste(centralp20$visitor_home_cbgs, centralp20$unique_ID, sep = "-")

centralp20<-centralp20%>%select(-c("placekey","parent_safegraph_place_id_original","parent_safegraph_place_id_new"))

centralp20<-centralp20%>%select(-c("parent_placekey"))

centralp19 <- left_join(centralp19, travel_19, by = "home_park_idn")

centralp20 <- left_join(centralp20, travel_20, by = "home_park_idn")

write.csv(centralp19,"/Users/wweng/Dropbox/Research/Pandemic and Recreation/Submission/PLOS ONE/R&R/Code/centralpark_visit_2019.csv")

write.csv(centralp20,"/Users/wweng/Dropbox/Research/Pandemic and Recreation/Submission/PLOS ONE/R&R/Code/centralpark_visit_2020.csv")


```

Import data
```{r}
centralpark2019<-read.csv("/Users/wweng/Dropbox/Research/Pandemic and Recreation/Submission/PLOS ONE/R&R/Code/centralpark_visit_2019.csv")

centralpark2019$year<-"2019"

centralpark2020 <- read.csv("/Users/wweng/Dropbox/Research/Pandemic and Recreation/Submission/PLOS ONE/R&R/Code/centralpark_visit_2020.csv")

centralpark2020$year<-"2020"

centralpark2019$id.x<-NULL
centralpark2019$park_id<-NULL
centralpark2019$id.y<-NULL

#combine data for 2019 and 2020

centralpark1920<-rbind(centralpark2019,centralpark2020)

# filter data within New York City
centralpark1920$visitorNYCity<-ifelse(substr(centralpark1920$visitor_home_cbgs.x,3,5)=="005"|substr(centralpark1920$visitor_home_cbgs.x,3,5)=="081"|substr(centralpark1920$visitor_home_cbgs.x,3,5)=="085"|substr(centralpark1920$visitor_home_cbgs.x,3,5)=="061"|substr(centralpark1920$visitor_home_cbgs.x,3,5)=="047",1,0)  

cp_NY<-centralpark1920%>%filter(visitorNYCity==1)

cp_NY$date <- paste(cp_NY$date_range_start,cp_NY$date_range_end,cp_NY$week)

```

Calculate travel cost
```{r}
#get blockgroup level income
#per capita income
bgincome_2019<-get_acs(geography = "block group",variables="B19301_001",year=2019,survey="acs5",state="NY")

bgincome_2020<-get_acs(geography = "block group",variables="B19301_001",year=2020,survey="acs5",state="NY")

bgincome_2019<-bgincome_2019%>%rename("percapitaincome"="estimate")
bgincome_2020<-bgincome_2020%>%rename("percapitaincome"="estimate")

cp_NY_2019<-cp_NY%>%filter(year==2019)
cp_NY_2020<-cp_NY%>%filter(year==2020)

cp_NY_2019<-cp_NY_2019%>%rename("GEOID"="visitor_home_cbgs.x")
cp_NY_2020<-cp_NY_2020%>%rename("GEOID"="visitor_home_cbgs.x")

cp_NY_2019$GEOID<-as.character(cp_NY_2019$GEOID)
cp_NY_2020$GEOID<-as.character(cp_NY_2020$GEOID)

cp_NY_2019<-cp_NY_2019%>%left_join(bgincome_2019,by="GEOID")
cp_NY_2020<-cp_NY_2020%>%left_join(bgincome_2020,by="GEOID")

cp_NY1920<-rbind(cp_NY_2019,cp_NY_2020)

#calculate travel cost
#calculate driving cost based on new york city taxi fare. On average, taxi fare is $2.5 for 321.90 meters, then $0.50 for every 321.90 meters

cp_NY1920$travelgroup<-ifelse(cp_NY1920$travel_meters<800.1,"0-800", ifelse(cp_NY1920$travel_meters<1600.1 & cp_NY1920$travel_meters>800,"800-1600",ifelse(cp_NY1920$travel_meters<2400.1 & cp_NY1920$travel_meters>1600,"1600-2400",ifelse(cp_NY1920$travel_meters<3200 & cp_NY1920$travel_meters>2400,"2400-3200",">3200"))))

cp_NY1920$taxicost<-ifelse(cp_NY1920$travelgroup=="0-800",0,(2.5+(cp_NY1920$travel_meters-321.90)/321.90*0.5)*2)

cp_NY1920$timecost<-cp_NY1920$percapitaincome/(2080*60)*cp_NY1920$travel_minutes*2 #on average, there are 2080 woring hours per year
  
cp_NY1920$totalcost<-cp_NY1920$taxicost+cp_NY1920$timecost

#export data
write.csv(cp_NY1920,"/Users/wweng/Dropbox/Research/Pandemic and Recreation/Submission/PLOS ONE/R&R/Code/centralpark_visit_1920.csv")
```

Summarize results by travelgroup
```{r}
travel_summarize<-cp_NY1920%>%
  group_by(travelgroup,year)%>%
  summarize(mean_cost=mean(totalcost,na.rm=TRUE),mean_taxicost=mean(taxicost,na.rm=TRUE),mean_timecost=mean(timecost,na.rm=TRUE))
```



